The Ontological Modal Collapse \\ as a Collapse of the Square of Opposition

### 2014-09-08 Papers/Universalis/ModalCollapse.tex
Boldify names of axioms and theorems

\begin{figure}[t]
\noindent \framebox[\columnwidth][r]{
\begin{minipage}{.94\columnwidth}\small
\begin{itemize}
\item[A1] Either a property or its negation is positive, but not
  both:
  $$\hol{\allq \varphi [P(\neg \varphi) \biimp \neg P(\varphi)]}$$ 
\item[A2] A property necessarily implied by a
  positive property is positive:
  $$\hol{\allq \varphi \allq \psi [(P(\varphi) \wedge \nec \allq x [\varphi(x)
  \imp \psi(x)]) \imp P(\psi)]}$$
\item[T1] Positive properties are possibly exemplified: 
  $$\hol{\allq \varphi [P(\varphi) \imp \pos \exq x \varphi(x)]}$$ 
\item[D1] A \emph{God-like} being possesses all positive properties: 
  $$\hol{G(x) \equiv \forall \varphi [P(\varphi) \imp \varphi(x)]}$$ 
\item[A3]  The property of being God-like is positive: 
  $$\hol{P(G)}$$
\item[C\phantom{1}] Possibly, a God-like being exists: $$\hol{\pos \exq x G(x)}$$
\item[A4]  Positive properties are necessarily positive: 
  $$\hol{\allq \varphi [P(\varphi) \imp \Box \; P(\varphi)]}$$ 
\item[D2] An \emph{essence} of an individual is a property possessed by it and necessarily implying any of its properties: $$\hol{\ess{\varphi}{x} \equiv \varphi(x) \wedge \allq
  \psi (\psi(x) \imp \nec \allq y (\varphi(y) \imp \psi(y)))}$$ 
\item[T2]  Being God-like is an essence of any
  God-like being: $$\hol{\allq x [G(x) \imp \ess{G}{x}]}$$
\item[D3] \emph{Necessary existence} of an individual is the necessary exemplification of all its essences: 
  $$\hol{\NE(x) \equiv \allq \varphi [\ess{\varphi}{x} \imp \nec
  \exq y \varphi(y)]}$$
\item[A5] Necessary existence is a positive property: $$\hol{P(\NE)}$$ 
\item[L1] If a god-like being exists, then necessarily a god-like being exists: 
  $$\hol{\exq x G(x) \imp \nec \exq y G(y)}$$
\item[L2] If possibly a god-like being exists, then necessarily a god-like being exists: 
  $$\hol{\pos \exq x G(x) \imp \nec \exq y G(y)} $$
%
\item[T3] Necessarily, a God-like being exists: $$\hol{\nec \exq x G(x)}$$ 
\end{itemize}
\end{minipage}
} \vskip-.5em
\caption{Scott's version of G\"odel's ontological argument \cite{ScottNotes}.\label{fig:scott}} 
\end{figure}

@@

A crucial step of most ontological arguments is the claim that 
if God's existence is possible, then it is necessary. 
This is Lemma L2 in G\"odel's proof. 
In the modal square of opposition (Fig. \ref{fig:square}), 
this is an unusual situation in which the \textbf{I} corner 
must imply and entail the \textbf{A} corner, 
in the particular case when $\phi$ is $\exq x G(x)$. 
G\"odel's proof shows that his axioms are strong enough 
to invert the direction of entailment for the sentence at issue. 
The question, however, is whether the axioms are not too strong, 
also allowing the inverted entailment for arbitrary $\phi$. 
That is essentially the question asked by Sobel \cite{Sobel1987}; 
and his proof of the modal collapse (MC) provides an affirmative 
answer. It is possible to show that this form of the modal 
collapse entails (in modal logic K) a collapse of the modal 
square (MC'), causing the subcontraries to entail (and even imply) their respective contraries. Normally, as shown in Fig. \ref{fig:square} in the modal square of opposition, only the other direction of entailment holds: the contraries entail their subcontraries, assuming the \emph{modal existential import} ExImp.

Furthermore, in any modal logic where the 
axiom T holds (i.e. where the accessibility relation is reflexive), a total collapse of the modalities (MC'') occurs. In particular, under this stronger form of modal collapse, the contraries entail their subcontraries even without the existential import.

@@

ToDo: include figure of Modal Square of Opposition.

\begin{figure}[t]
\noindent \framebox[\columnwidth][r]{
\begin{minipage}{.94\columnwidth}\small
\begin{itemize}
\item[MC] Everything that is the case is so necessarily:
  $$\hol{\allq \phi [\phi \imp \nec \phi]}$$ 
\item[MC'] Everything that is possible is necessary:
  $$\hol{\allq \phi [\pos \phi \imp \nec \phi]}$$ 
\item[T] Everything that is necessary is the case:
  $$\hol{\allq \phi [\nec \phi \imp \phi]}$$ 
\item[ExImp] (Modal Existential Import):
  $$\hol{\pos \top}$$ 
\item[AI] Everything that is necessary is possible:
  $$\hol{\allq \phi [\nec \phi \imp \pos \phi]}$$ 
\item[MC''] Modalities collapse completely:
  $$\hol{\allq \phi [\phi \biimp \pos \phi \biimp \nec \phi]}$$
\end{itemize}
\end{minipage}
} \vskip-.5em
\caption{Modal Collapse \cite{todo:Sobel1987,SobelBook}.\label{fig:collapse}} 
\end{figure}

\begin{Huge}

$$
\nec \ \phi
$$

$$
\pos \ \phi
$$

$$
\neg \ \nec \ \phi
$$

$$
\neg \ \pos \ \phi
$$

$$
\pos \ \exists x \ G(x)
$$

$$
\nec \ \exists x \ G(x)
$$

\end{Huge}

>>>>

\begin{figure}[t]
\noindent \framebox[\columnwidth][r]{
\begin{minipage}{.94\columnwidth}\small
\begin{itemize}
\item[\textbf{A1}] Either a property or its negation is positive, but not
  both:
  $$\hol{\allq \varphi [P(\neg \varphi) \biimp \neg P(\varphi)]}$$ 
\item[\textbf{A2}] A property necessarily implied by a
  positive property is positive:
  $$\hol{\allq \varphi \allq \psi [(P(\varphi) \wedge \nec \allq x [\varphi(x)
  \imp \psi(x)]) \imp P(\psi)]}$$
\item[\textbf{T1}] Positive properties are possibly exemplified: 
  $$\hol{\allq \varphi [P(\varphi) \imp \pos \exq x \varphi(x)]}$$ 
\item[\textbf{D1}] A \emph{God-like} being possesses all positive properties: 
  $$\hol{G(x) \equiv \forall \varphi [P(\varphi) \imp \varphi(x)]}$$ 
\item[\textbf{A3}]  The property of being God-like is positive: 
  $$\hol{P(G)}$$
\item[\textbf{C\phantom{1}}] Possibly, a God-like being exists: $$\hol{\pos \exq x G(x)}$$
\item[\textbf{A4}]  Positive properties are necessarily positive: 
  $$\hol{\allq \varphi [P(\varphi) \imp \Box \; P(\varphi)]}$$ 
\item[\textbf{D2}] An \emph{essence} of an individual is a property possessed by it and necessarily implying any of its properties: $$\hol{\ess{\varphi}{x} \equiv \varphi(x) \wedge \allq
  \psi (\psi(x) \imp \nec \allq y (\varphi(y) \imp \psi(y)))}$$ 
\item[\textbf{T2}]  Being God-like is an essence of any
  God-like being: $$\hol{\allq x [G(x) \imp \ess{G}{x}]}$$
\item[\textbf{D3}] \emph{Necessary existence} of an individual is the necessary exemplification of all its essences: 
  $$\hol{\NE(x) \equiv \allq \varphi [\ess{\varphi}{x} \imp \nec
  \exq y \varphi(y)]}$$
\item[\textbf{A5}] Necessary existence is a positive property: $$\hol{P(\NE)}$$ 
\item[\textbf{L1}] If a god-like being exists, then necessarily a god-like being exists: 
  $$\hol{\exq x G(x) \imp \nec \exq y G(y)}$$
\item[\textbf{L2}] If possibly a god-like being exists, then necessarily a god-like being exists: 
  $$\hol{\pos \exq x G(x) \imp \nec \exq y G(y)} $$
%
\item[\textbf{T3}] Necessarily, a God-like being exists: $$\hol{\nec \exq x G(x)}$$ 
\end{itemize}
\end{minipage}
} \vskip-.5em
\caption{Scott's version of G\"odel's ontological argument \cite{ScottNotes}.\label{fig:scott}} 
\end{figure}

@@

A crucial step of most ontological arguments is the claim that 
if God's existence is possible, then it is necessary. 
This is Lemma \textbf{L2} in G\"odel's proof. 
In the modal square of opposition (Fig. \ref{fig:square}), 
this is an unusual situation in which the \textbf{I} corner 
must imply and entail the \textbf{A} corner, 
in the particular case when $\phi$ is $\exq x G(x)$. 
G\"odel's proof shows that his axioms are strong enough 
to invert the direction of entailment for the sentence at issue. 
The question, however, is whether the axioms are not too strong, 
also allowing the inverted entailment for arbitrary $\phi$. 
That is essentially the question asked by Sobel \cite{Sobel1987}; 
and his proof of the modal collapse (\textbf{MC}) provides an affirmative 
answer. It is possible to show that this form of the modal 
collapse entails (in modal logic \textbf{\emph{K}}) a collapse of the modal 
square (\textbf{MC'}), causing the subcontraries to entail (and even imply) their respective contraries. Normally, as shown in Fig. \ref{fig:square}, in the modal square of opposition only the other direction of entailment holds: the contraries entail their subcontraries, assuming the \emph{modal existential import} \textbf{ExImp}.

Moreover, in any modal logic where the 
axiom \textbf{T} holds (i.e. where the accessibility relation is reflexive), a total collapse of the modalities (\textbf{MC''}) occurs. Interestingly, under this stronger form of modal collapse, the contraries entail their subcontraries even without the existential import.

@@

\begin{figure}[t]
\noindent \framebox[\columnwidth][r]{
\begin{minipage}{.94\columnwidth}\small
\begin{itemize}
\item[\textbf{MC}] Everything that is the case is so necessarily:
  $$\hol{\allq \phi [\phi \imp \nec \phi]}$$ 
\item[\textbf{MC'}] Everything that is possible is necessary:
  $$\hol{\allq \phi [\pos \phi \imp \nec \phi]}$$ 
\item[\textbf{T}] Everything that is necessary is the case:
  $$\hol{\allq \phi [\nec \phi \imp \phi]}$$ 
\item[\textbf{ExImp}] (Modal Existential Import):
  $$\hol{\pos \top}$$ 
\item[\textbf{AI}] Everything that is necessary is possible:
  $$\hol{\allq \phi [\nec \phi \imp \pos \phi]}$$ 
\item[\textbf{MC''}] Modalities collapse completely:
  $$\hol{\allq \phi [\phi \biimp \pos \phi \biimp \nec \phi]}$$
\end{itemize}
\end{minipage}
} \vskip-.5em
\caption{Modal Collapse}
\label{fig:collapse} 
\end{figure}

\newcommand{\collapse}[1]{\mathit{collapse}(#1)}

The

Let $collapse{\phi} \equiv \phi \imp \nec \phi$. It is known that  Variants of the ontological argument
